Finch: Sparse and Structured Array Programming with Control Flow

### 2024-03-29 sections/tensors.tex
Update on Overleaf.

\begin{figure}[htbp]
    \scriptsize
    \raggedright
\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors
of size $dims$ and an initial value of $init$. Requires the level to be in
read-only mode.
\paragraph{$freeze(lvl)$} Finalizes the updates in the level, and readies the
level for reading. Requires the level to be in update-only mode.
\paragraph{$thaw(lvl)$} Prepares the level to accept updates. Requires the level
to be in read-only mode.
\paragraph{$unfurl(fiber(lvl, pos), ext, mode)$} Unfurls the fiber at position
$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
returns a looplet nest over the values in the read-only fiber.  When $mode =
\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
fiber. Often, skipping over mutable locations allows the level to know which
locations must be stored. Additionally, a dirty bit may be used to communicate
whether the mutable subfiber has been written to, which allows the parent fiber
to know whether the subfiber must be stored explicitly.
\paragraph{$assemble(lvl, pos_{start}, pos_{stop})$} Allocates subfibers in the
level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
only ever called on unassembled positions, but some levels (such as dense levels
or bytemaps) may support reassembly.
\caption{The five functions that define a level.}
\end{figure}

@@

    \begin{wrapfigure}{r}{.35\textwidth}
        \centering
        \scriptsize
        \begin{tabular}{|c|c|c|c|l|}
            \hline
            \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singletons} & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\
            \hline
             &  &  &  & Dense \\
            \hline
             &  & \checkmark &  & DenseRLE \\
            \hline
            \checkmark &  &  &  & Sparse \\
            \hline
            \checkmark &  &  & \checkmark & SparsePinpoint \\
            \hline
            \checkmark &  & \checkmark &  & SparseRLE \\
            \hline
            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
            \hline
            \checkmark & \checkmark &  &  & SparseVBL \\
            \hline
            \checkmark & \checkmark &  & \checkmark & SparseBand \\
            \hline
        \end{tabular}
        \caption{All combinations of relevant structural properties and their
        corresponding formats.  Note that blocks and runs need not be considered
        together because we must store a run length for each run, and so there
        isn't a significant storage benefit to combining them. Blocks and
        singletons only make sense in the contex of sparsity, so we don't
        consider them together either. We omit such combinations
        from the exhaustive table.}
        \label{tab:TypesOfStructure}
    \end{wrapfigure}

@@

\begin{table}[ht]
    \centering
    \scriptsize
    \begin{tabular}{p{14cm}}
    \hline
    \textbf{Sequential Reading and Updating} \\
    \hline
    \textbf{Dense}:
    The dense format is the simplest format, mapping $fiber(l, p)[i] \rightarrow fiber(l.lvl, p \times l.shape + i)$.
    %
    This format is used to store dense data and is often a convenient format for the root level of a tensor.
    %
    Due to its simplicity, freezing and thawing the level are no-ops. \\
    \textbf{DenseRLE}:
    Used to represent runs of repeated values, storing two vectors, $right$ and $ptr$, with $qth$ run in the $p^th$ subfiber starting and ending at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
    %
    A challenge arises for this level: it is difficult to merge duplicate runs.
    %
    Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
    %
    Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
    %
    Instead, we merge the duplicates during the freeze phase.
    %
    We $freeze$ the sublevel, $declare$ a separate sublevel $buf$ as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer. \\
    %
    %It also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseList}:
    The simplest sparse format, used to construct popular formats like CSR, CSC, DCSR, DCSC, and CSF.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. \\
    \textbf{SparsePinpoint}:
    Similar to SparseList, but only one nonzero in each subfiber, eliminating the need for the $ptr$ field.
    %
    It stores a vector $idx$, such that $idx[p]$ is the nonzero index in the subfiber at position $p$. \\
    \textbf{SparseRLE}:
    Similar to DenseRLE level, but because runs are sparse, we must also store the start of each run.
    %
    It stores three vectors $left$, $right$, and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
    %
    Like DenseRLE, it also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseInterval}:
    Similar to SparseRLE, but only stores one run per subfiber, eliminating the need for the $ptr$ field.
    %
    This level does not deduplicate as it cannot store intermediate results with more than one run.
    %
    It stores two vectors, such that the run in subfiber $p$ begins and ends at $left[p]$ and $right[p]$ respectively. \\
    \textbf{SparseVBL}: Used to represent blocked data. It stores three vectors, $idx$, $ptr$, and $ofs$, such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ ending at index $idx[ptr[p] + q]$ in the subfiber at position $p$. \\
    \textbf{SparseBand}: Similar to SparseVBL, but stores only one block per subfiber, eliminating the need for the $ptr$ field. It stores two vectors $idx$ and $ofs$, such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block ending at $idx[p]$ in subfiber $p$. \\
    \hline
    \textbf{Random Access Reading and Updating} \\
    \hline
    \textbf{SparseHash}:
    The sparse hash format uses a hash table to store the locations of nonzeros, and sorts the unique indices for iteration during the freeze phase.
    %
    This allows for efficient random access, but not incremental construction, as the freeze phase runs in time proportional to the number of nonzeros in the entire level.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. 
    %
    Also stores a hash table $tbl$ for construction and random access in the level. \\
    \textbf{SparseBytemap}
    The SparseBytemap format uses a bytemap to store which locations have been written to.
    %
    Unlike the SparseHash format, the bytemap assembles the entire space of possible subfibers.
    %
    This accelerates random access in the format, but requires a high memory overhead.
    %
    Because we don't want to reallocate all of the memory in each iteration, the declaration of this format instead re-assembles only the dirty locations in the tensor.
    %
    This format is analogous to the default workspace format used by TACO.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$.
    %
    These vectors are used to collect dirty locations.
    %
    It also stores $tbl$, a dense array of booleans such that $tbl[shape * p + i]$ is true when there is a nonzero at index $i$ in the subfiber at position $p$. \\
    \hline
    \end{tabular}
    \caption{The main level formats supported by Finch. Note that all non-leaf
    levels store a the dimension of the subfibers and a child level. Since we
    must be able to handle the case where a sublevel is not stored because a
    parent level is sparse, all of Finch's sparse formats use a dirty bit during
    writing to determine whether the sublevel has been modified from it's
    default fill value and thus, whether it needs to be stored.}
    \label{table:formats}
\end{table}

>>>>

\begin{figure}[htbp]
    \footnotesize
    \raggedright
$declare(lvl, init, dims...)$: Declares the level to hold subtensors
of size $dims$ and an initial value of $init$. Requires the level to be in
read-only mode.

$freeze(lvl)$: Finalizes the updates in the level, and readies the
level for reading. Requires the level to be in update-only mode.

$thaw(lvl)$: Prepares the level to accept updates. Requires the level
to be in read-only mode.

$unfurl(fiber(lvl, pos), ext, mode)$: Unfurls the fiber at position
$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
returns a looplet nest over the values in the read-only fiber.  When $mode =
\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
fiber. Often, skipping over mutable locations allows the level to know which
locations must be stored. Additionally, a dirty bit may be used to communicate
whether the mutable subfiber has been written to, which allows the parent fiber
to know whether the subfiber must be stored explicitly.

$assemble(lvl, pos_{start}, pos_{stop})$: Allocates subfibers in the
level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
only ever called on unassembled positions, but some levels (such as dense levels
or bytemaps) may support reassembly.
\caption{The five functions that define a level.}
\end{figure}

@@

    \begin{wrapfigure}{r}{.35\textwidth}
        \centering
        \footnotesize
        \begin{tabular}{|c|c|c|c|l|}
            \hline
            \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singletons} & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\
            \hline
             &  &  &  & Dense \\
            \hline
             &  & \checkmark &  & DenseRLE \\
            \hline
            \checkmark &  &  &  & Sparse \\
            \hline
            \checkmark &  &  & \checkmark & SparsePinpoint \\
            \hline
            \checkmark &  & \checkmark &  & SparseRLE \\
            \hline
            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
            \hline
            \checkmark & \checkmark &  &  & SparseVBL \\
            \hline
            \checkmark & \checkmark &  & \checkmark & SparseBand \\
            \hline
        \end{tabular}
        \caption{All combinations of relevant structural properties and their
        corresponding formats.  Note that blocks and runs need not be considered
        together because we must store a run length for each run, and so there
        isn't a significant storage benefit to combining them. Blocks and
        singletons only make sense in the contex of sparsity, so we don't
        consider them together either. We omit such combinations
        from the exhaustive table.}
        \label{tab:TypesOfStructure}
    \end{wrapfigure}

@@

\begin{table}[ht]
    \centering
    \footnotesize
    \begin{tabular}{p{14cm}}

    \hline
    \textbf{Sequential Reading and Updating} \\
    \hline
    \textbf{Dense}:
    The dense format is the simplest format, mapping $fiber(l, p)[i] \rightarrow fiber(l.lvl, p \times l.shape + i)$.
    %
    This format is used to store dense data and is often a convenient format for the root level of a tensor.
    %
    Due to its simplicity, freezing and thawing the level are no-ops. \\
    \textbf{DenseRLE}:
    Used to represent runs of repeated values, storing two vectors, $right$ and $ptr$, with $qth$ run in the $p^th$ subfiber starting and ending at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
    %
    A challenge arises for this level: it is difficult to merge duplicate runs.
    %
    Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
    %
    Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
    %
    Instead, we merge the duplicates during the freeze phase.
    %
    We $freeze$ the sublevel, $declare$ a separate sublevel $buf$ as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer. \\
    %
    %It also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseList}:
    The simplest sparse format, used to construct popular formats like CSR, CSC, DCSR, DCSC, and CSF.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. \\
    \textbf{SparsePinpoint}:
    Similar to SparseList, but only one nonzero in each subfiber, eliminating the need for the $ptr$ field.
    %
    It stores a vector $idx$, such that $idx[p]$ is the nonzero index in the subfiber at position $p$. \\
    \textbf{SparseRLE}:
    Similar to DenseRLE level, but because runs are sparse, we must also store the start of each run.
    %
    It stores three vectors $left$, $right$, and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
    %
    Like DenseRLE, it also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseInterval}:
    Similar to SparseRLE, but only stores one run per subfiber, eliminating the need for the $ptr$ field.
    %
    This level does not deduplicate as it cannot store intermediate results with more than one run.
    %
    It stores two vectors, such that the run in subfiber $p$ begins and ends at $left[p]$ and $right[p]$ respectively. \\
    \textbf{SparseVBL}: Used to represent blocked data. It stores three vectors, $idx$, $ptr$, and $ofs$, such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ ending at index $idx[ptr[p] + q]$ in the subfiber at position $p$. \\
    \textbf{SparseBand}: Similar to SparseVBL, but stores only one block per subfiber, eliminating the need for the $ptr$ field. It stores two vectors $idx$ and $ofs$, such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block ending at $idx[p]$ in subfiber $p$. \\

    \hline
    \textbf{Random Access Reading and Updating} \\
    \hline
    \textbf{SparseHash}:
    The sparse hash format uses a hash table to store the locations of nonzeros, and sorts the unique indices for iteration during the freeze phase.
    %
    This allows for efficient random access, but not incremental construction, as the freeze phase runs in time proportional to the number of nonzeros in the entire level.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. 
    %
    Also stores a hash table $tbl$ for construction and random access in the level. \\
    \textbf{SparseBytemap}
    The SparseBytemap format uses a bytemap to store which locations have been written to.
    %
    Unlike the SparseHash format, the bytemap assembles the entire space of possible subfibers.
    %
    This accelerates random access in the format, but requires a high memory overhead.
    %
    Because we don't want to reallocate all of the memory in each iteration, the declaration of this format instead re-assembles only the dirty locations in the tensor.
    %
    This format is analogous to the default workspace format used by TACO.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$.
    %
    These vectors are used to collect dirty locations.
    %
    It also stores $tbl$, a dense array of booleans such that $tbl[shape * p + i]$ is true when there is a nonzero at index $i$ in the subfiber at position $p$. \\
    \hline
    \textbf{Leaf Levels} \\
    \hline
    \textbf{Element}:
    The element level uses an array $val$ to store a value for each position $p$. The zero (fill) value is configurable.\\
    \textbf{Pattern}:
    The pattern level statically represents a leaf level with a fill value of $false$ and whose stored values are all $true$. \\
    \hline
    \end{tabular}
    \caption{The main level formats supported by Finch. Note that all non-leaf
    levels store a the dimension of the subfibers and a child level. Since we
    must be able to handle the case where a sublevel is not stored because a
    parent level is sparse, all of Finch's sparse formats use a dirty bit during
    writing to determine whether the sublevel has been modified from it's
    default fill value and thus, whether it needs to be stored.}
    \label{table:formats}
\end{table}
